In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-04 paper/main.tex
Address point about PVS permeability.

Human and rodent observations indicate that tracers concentrate in perivascular spaces surrounding the pial and subarachnoid vasculature \cite{zhang1990interrelationships,zhang1992directional, bedussi2017paravascular, mestre2018flow, bedussi2018paravascular, eide2024functional}. However, it remains unclear whether such enrichment patterns necessitate a structural barrier, such as a membrane with limited permeability, or if the patterns could result from enhanced flow or mixing in these areas alone. We therefore next investigate how the permeability of the interface between the PVS and the surrounding CSF affects tracer enrichment around the major arterial trunks (\Cref{fig:compartmentalization}A--B). To this end, we compare models with high and low  permeabilities, representing a highly permeable PVS-CSF interface (functional compartmentalization) and a less permeable interface (structural compartmentalization), respectively. For the low permeability model, we set the permeability to $3.8 \cdot 10^{-7}\,$m/s, consistent with previous estimates for the endfoot sheath surrounding penetrating arterioles \cite{koch2023estimates}, which we consider to be a lower bound for the surface PVS-CSF permeability. For the high permeability model, we increase this permeability by a factor of 100. For both scenarios, we consider the high PVS flow regime as examined in the previous section.

@@

% Compartmentalization of the PVS
To what extent are surface PVSs in communication with the surrounding
SAS and to what extent are these structurally separated compartments?
Originally, Weller and coauthors~\cite{zhang1990interrelationships,
  zhang1992directional, weller2005microscopic} identified thin sheaths
of pial cells surrounding human surface arteries and penetrating
arterioles (but not veins or venules). On the other hand, Bakker and
coauthors report that the subarachnoid space, the cisterns, ventricles
and penetrating periarteriolar spaces form a continuous CSF
space~\cite{bedussi2017paravascular}. Thorne and colleagues study the
architectures of rat perivascular spaces in detail, emphasizing the
presence of openings (stomata) on the interface between the
vasculature and the CSF within the SAS~\cite{pizzo2018intrathecal,
  hannocks2018molecular}. Further, Mestre et
al~\cite{mestre2022periarteriolar} study the properties of pial
perivascular spaces in mice, and report of pial cells forming sheaths
for larger surface arteries and partially cover smaller surface
arteries, with higher coverage in ventral SAS regions. Our results
demonstrate that the PVSs may act as rapid transport pathways even
with only a partial barrier between the PVS and surrounding CSF, thus
contributing to quantifying the properties of the human PVS-CSF
interface~\cite{eide2024functional}.

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs_a}C). The corresponding contribution to the periarterial
flow velocity is defined for each segment $\Lambda_i$ as $\langle Q'_i
\rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the cross-section
area of the PVS segment and $\langle Q_i' \rangle$ is the estimated
mean flow rate (see Supplementary information S1.4 for more details). Two
wall motion patterns are considered, yielding $\hat{u}_a^c$ and
$\hat{u}_a^v$ respectively: \emph{cardiac} pulsations with $f =
1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$ (\Cref{fig:pvs_a}D), and very low-frequency \emph{vasomotion} with $f = 0.1$Hz, $\lambda =
0.02$m, and $\varepsilon = 10\%$ (\Cref{fig:pvs_a}E). No flow induced by peristalsis is considered for the perivenous network. In the baseline model, we combine the contributions from CSF production and cardiac peristalsis to set $\hat{u}$ in~\eqref{eq:multi_transport} as
$\hat{u}_a = \hat{u}_a^p + \hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v
= \hat{u}_v^p$ in $\Lambda_v$.

@@

    respectively, but with dilated PVS.}

>>>>

Human and rodent observations indicate that tracers concentrate in perivascular spaces surrounding the pial and subarachnoid vasculature \cite{zhang1990interrelationships,zhang1992directional, bedussi2017paravascular, mestre2018flow, bedussi2018paravascular, eide2024functional}. However, it remains unclear whether such enrichment patterns necessitate a structural barrier, such as a membrane with limited permeability, or if the patterns could result from enhanced flow or mixing in these areas alone. We therefore next investigate how the permeability of the interface between the PVS and the surrounding CSF affects tracer enrichment around the major arterial trunks (\Cref{fig:compartmentalization}A--B). To this end, we compare models with high and low  permeabilities, representing a highly permeable PVS-CSF interface (functional compartmentalization) and a less permeable interface (structural compartmentalization), respectively. For the low permeability model, we set the permeability to $3.8 \cdot 10^{-7}\,$m/s, consistent with previous estimates for the endfoot sheath surrounding penetrating arterioles \cite{koch2023estimates}, which we consider to be a lower bound for the surface PVS-CSF permeability. For the high permeability model, we increase this permeability by a factor of 100. We emphasize that, for both scenarios, we consider the high PVS flow regime as examined in the previous section.

@@

% Compartmentalization of the PVS
To what extent are surface PVSs in communication with the surrounding
SAS and to what extent are these structurally separated compartments?
Originally, Weller and coauthors~\cite{zhang1990interrelationships,
  zhang1992directional, weller2005microscopic} identified thin sheaths
of pial cells surrounding human surface arteries and penetrating
arterioles (but not veins or venules). On the other hand, Bakker and
coauthors report that the subarachnoid space, the cisterns, ventricles
and penetrating periarteriolar spaces form a continuous CSF
space~\cite{bedussi2017paravascular}. Thorne and colleagues study the
architectures of rat perivascular spaces in detail, emphasizing the
presence of openings (stomata) on the interface between the
vasculature and the CSF within the SAS~\cite{pizzo2018intrathecal,
  hannocks2018molecular}. Further, Mestre et
al~\cite{mestre2022periarteriolar} study the properties of pial
perivascular spaces in mice, and report of pial cells forming sheaths
for larger surface arteries and partially cover smaller surface
arteries, with higher coverage in ventral SAS regions. Our results
demonstrate that the PVSs may act as rapid transport pathways even
with only a partial barrier between the PVS and surrounding CSF, thus
contributing to quantifying the properties of the human PVS-CSF
interface~\cite{eide2024functional}. In this regard, we emphasize that
we report on the isolated effect of variations in permeability on
molecular transport, separated from any effects on fluid flow in the
PVS. Clearly, one may expect that the presence and/or properties of
such a barrier would affect also PVS flow. However, considering the
uncertainty associated with PVS flow characteristics in the human SAS
in general and the lack of computational models describing net PVS
flow in the presence of fluid exchange, this remains challenging to
quantify.

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs_a}C). The corresponding contribution to the
periarterial flow velocity is defined for each segment $\Lambda_i$ as
$\langle Q'_i \rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the
cross-section area of the PVS segment and $\langle Q_i' \rangle$ is
the estimated mean flow rate (see Supplementary information S1.4 for
more details). Two wall motion patterns are considered, yielding
$\hat{u}_a^c$ and $\hat{u}_a^v$ respectively: \emph{cardiac}
pulsations with $f = 1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$
(\Cref{fig:pvs_a}D), and very low-frequency \emph{vasomotion} with $f
= 0.1$Hz, $\lambda = 0.02$m, and $\varepsilon = 10\%$
(\Cref{fig:pvs_a}E). No flow induced by peristalsis is considered for
the perivenous network. In the baseline model, we combine the
contributions from CSF production and cardiac peristalsis to set
$\hat{u}$ in~\eqref{eq:multi_transport} as $\hat{u}_a = \hat{u}_a^p +
\hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v = \hat{u}_v^p$ in
$\Lambda_v$. We remark that none of these models account for fluid
exchange across the lateral outer interface of the PVSs (over
$\Gamma_a, \Gamma_v$), while fluid entry and exit is allowed at the
end nodes of the PVS networks.

@@

    respectively, but with dilated PVS. }
